---
title: "Assignment 2"
output: html_document
date: "2026-10-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
UniversalBank <- read.csv("C:/Users/sophi/Downloads/UniversalBank (1).csv")
summary(UniversalBank)
library(caret)
library(class)
library(gmodels)

# Removing ID and ZIP Code
UniversalBank <- UniversalBank[, -c(1,5)]
# Checking remaining variables
names(UniversalBank)


# Telling R that Education is categorical
UniversalBank$Education <- as.factor(UniversalBank$Education)
class(UniversalBank$Education)

# Creating dummy variables for Education
dummy_model <- dummyVars(~Education, data = UniversalBank)

# Applying the dummy-variable model
Education_dummy <- predict(dummy_model, UniversalBank)

# Looking at the first rows
head(Education_dummy)


# Removing the original Education variable
UniversalBank <- UniversalBank[, -6]

# Adding the three Education dummy variables
UniversalBank <- cbind(UniversalBank, Education_dummy)

# Checking the variables
names(UniversalBank)


UniversalBank$Personal.Loan <- as.factor(UniversalBank$Personal.Loan)
class(UniversalBank$Personal.Loan)

# Selecting 60% of the data for training
set.seed(123)

train_rows <- createDataPartition(
  UniversalBank$Personal.Loan,
  p = 0.6,
  list = FALSE
)

# Creating the training dataset
Train <- UniversalBank[train_rows, ]

# Creating the validation dataset
Validation <- UniversalBank[-train_rows, ]

# Separating the training predictors
Train_Predictors <- Train[, -7]

# Separating the validation predictors
Validation_Predictors <- Validation[, -7]

# Saving the training class labels
Train_labels <- Train$Personal.Loan

# Saving the validation class labels
Validation_labels <- Validation$Personal.Loan

# Creating the normalization model
norm_model <- preProcess(
  Train_Predictors,
  method = "range"
)

# Normalizing the training predictors
Train_Predictors <- predict(
  norm_model,
  Train_Predictors
)

# Normalizing the validation predictors
Validation_Predictors <- predict(
  norm_model,
  Validation_Predictors
)

# Checking the normalized training data
head(Train_Predictors)



# Question 1
# Creating the new customer
New_Customer <- data.frame(
  Age = 40,
  Experience = 10,
  Income = 84,
  Family = 2,
  CCAvg = 2,
  Mortgage = 0,
  Securities.Account = 0,
  CD.Account = 0,
  Online = 1,
  CreditCard = 1,
  Education.1 = 0,
  Education.2 = 1,
  Education.3 = 0
)

# Normalizing the new customer
New_Customer_normalized <- predict(
  norm_model,
  New_Customer
)


# Classifying the new customer with k = 1
Customer_k1 <- knn(
  Train_Predictors,
  New_Customer_normalized,
  cl = Train_labels,
  k = 1
)

# Showing the predicted class
Customer_k1

```
Using k-NN with k=1 the customer is classified as class 0. So the model predicts that the customer will not accept the personal loan offer.

```{r}
#Question 2:
# Setting the k values to test
Search_grid <- expand.grid(
  k = c(1, 3, 5, 7, 9, 11, 13, 15, 17, 19)
)

# Tuning the kNN model
set.seed(123)

model_k <- train(
  x = Train_Predictors,
  y = Train_labels,
  method = "knn",
  tuneGrid = Search_grid
)

# Showing the results
model_k

# Predicting the validation data using the selected k
Validation_Predictions <- knn(
  Train_Predictors,
  Validation_Predictors,
  cl = Train_labels,
  k = 1
)

# Showing the first predictions
head(Validation_Predictions)

# Comparing the predicted and actual validation classes
CrossTable(
  Validation_labels,
  Validation_Predictions,
  prop.chisq = FALSE
)

```
The selected value is k=1. The train function compared the different values of k and selected k=1 because it had the highest accuracy, about 95.19%. Although larger values of k provide more smoothing, their accuracy was lower in this case, so I used k=1 for the k-NN model.

```{r}
#QUESTION 3
```
Using the best value k=1, the confusion matrix for the validation data shows 1,787 true negatives, 21 false positives, 62 false negatives, and 130 true positives. Overall, 1,917 out of 2,000 observations were correctly classified, giving an accuracy of 95.85%.
```{r}

#QUESTION 4
# Classifying the new customer using the best k
Customer_best_k <- knn(
  Train_Predictors,
  New_Customer_normalized,
  cl = Train_labels,
  k = 1
)

# Showing the predicted class
Customer_best_k

```
Using the best value k=1, the customer is classified as class 0. So the model predicts that this customer will not accept the personal loan offer.
```{r}


#QUESTION 5
# Selecting 50% of the data for training
set.seed(123)

train_rows2 <- createDataPartition(
  UniversalBank$Personal.Loan,
  p = 0.5,
  list = FALSE
)

# Creating the training dataset
Train2 <- UniversalBank[train_rows2, ]

# Creating the remaining dataset
Remaining <- UniversalBank[-train_rows2, ]

# Selecting 60% of the remaining data for validation
validation_rows2 <- createDataPartition(
  Remaining$Personal.Loan,
  p = 0.6,
  list = FALSE
)

# Creating the validation dataset
Validation2 <- Remaining[validation_rows2, ]

# Creating the test dataset
Test2 <- Remaining[-validation_rows2, ]

# Checking the size of each dataset
nrow(Train2)
nrow(Validation2)
nrow(Test2)

# Separating the training predictors
Train2_Predictors <- Train2[, -7]

# Separating the validation predictors
Validation2_Predictors <- Validation2[, -7]

# Separating the test predictors
Test2_Predictors <- Test2[, -7]

# Saving the training class labels
Train2_labels <- Train2$Personal.Loan

# Saving the validation class labels
Validation2_labels <- Validation2$Personal.Loan

# Saving the test class labels
Test2_labels <- Test2$Personal.Loan

# Creating the new normalization model
norm_model2 <- preProcess(
  Train2_Predictors,
  method = "range"
)

# Normalizing the training predictors
Train2_Predictors <- predict(
  norm_model2,
  Train2_Predictors
)

# Normalizing the validation predictors
Validation2_Predictors <- predict(
  norm_model2,
  Validation2_Predictors
)

# Normalizing the test predictors
Test2_Predictors <- predict(
  norm_model2,
  Test2_Predictors
)

# Predicting the training data
Train2_Predictions <- knn(
  Train2_Predictors,
  Train2_Predictors,
  cl = Train2_labels,
  k = 1
)

# Predicting the validation data
Validation2_Predictions <- knn(
  Train2_Predictors,
  Validation2_Predictors,
  cl = Train2_labels,
  k = 1
)

# Predicting the test data
Test2_Predictions <- knn(
  Train2_Predictors,
  Test2_Predictors,
  cl = Train2_labels,
  k = 1
)

# Creating the training confusion matrix
CrossTable(
  Train2_labels,
  Train2_Predictions,
  prop.chisq = FALSE
)

# Creating the validation confusion matrix
CrossTable(
  Validation2_labels,
  Validation2_Predictions,
  prop.chisq = FALSE
)

# Creating the test confusion matrix
CrossTable(
  Test2_labels,
  Test2_Predictions,
  prop.chisq = FALSE
)

```
After splitting the data, I had 2,500 observations for training, 1,500 for validation, and 1,000 for testing. With k=1, the training accuracy was 100 percent, while the validation and test accuracies were 96% and 95.6%.
The training accuracy is higher because the model is classifying the same observations that were used in training. The validation and test results are a little lower, but they are very close to each other, which shows that the model gives similar results on data it did not use for training.
But the model is better at predicting class 0 than class 1. In the test set, it correctly predicted 63 out of 96 customers who actually accepted the loan, which is about 65.6%. Since only 9.6% of customers accepted the loan, the data are unbalanced. The overall accuracy is high partly because most customers did not accept the loan, so accuracy alone can hide the fact that the model is less effective at identifying the smaller group of customers who actually accepted it.
